Competing Bandits: \\ The Perils of Exploration under Competition

### 2019-08-18 journal paper/content/sec-model.tex
first pass through intro and model section

The information structure is as follows. The prior $\prior$ is known to everyone. The mean rewards $\mu_a$ are not revealed to anybody. Each agent knows both principals' algorithms, and the global round when (s)he arrives, \emph{but not} the rewards of the previous agents. Each principal is completely unaware of the rounds when the other is chosen.

\xhdr{Some terminology.} The two principals are called ``Principal
1" and ``Principal 2".
%Principal $i\in \{1,2\}$ is also denoted \prin[i].
The algorithm of principal $i\in \{1,2\}$ is called ``algorithm $i$" and denoted
\alg[i]. The agent in global round $t$ is called ``agent $t$"; the
chosen principal is denoted $i_t$.

@@

\xhdr{Agents' response.}
Each agent $t$ chooses principal $i_t$ as as follows: it chooses a distribution over the principals, and then draws independently from this distribution. Let $p_t$ be the probability of choosing principal $1$ according to this distribution. Below we specify $p_t$; we need to be careful so as to avoid a circular definition.

@@

\xhdr{Response functions.}
We use the response function $\respF$ to characterize the amount of rationality and competitiveness in our model. We assume that $\respF$ is monotonically non-decreasing, is larger than $\tfrac12$ on the interval $(0,1]$, and smaller than $\tfrac12$ on the interval $[-1,0)$. Beyond that, we consider three specific models, listed in the order of decreasing rationality and competitiveness (see \reffig{fig:response-functions}):

@@

We will emphasize several specific algorithms or classes thereof:

\begin{itemize}
\item ``smart" MAB algorithms that combine exploration and exploitation, such as UCB1 \cite{bandits-ucb1} and Successive Elimination \cite{EvenDar-icml06}. These algorithms achieve
        $\BIR(n) \leq  \tilde{O}(n^{-1/2})$
    for all priors and all (or all but a very few) steps $n$. This bound is known to be tight for any fixed $n$.
    \footnote{This follows from the lower-bound analysis in \cite{bandits-exp3}.}

@@

We focus on MAB algorithms such that $\BIR(n)$ is non-increasing; we call such algorithms \emph{monotone}. While some reasonable MAB algorithms may occasionally violate monotonicity, they can usually be easily modified so that monotonicity violations either vanish altogether, or only occur at very specific rounds (so that agents are extremely unlikely to exploit them in practice).

@@

@@

>>>>

\gaedit{
The information structure is as follows. The prior $\prior$ is known to everyone. The mean rewards $\mu_a$ are not revealed to anybody and each principal is completely unaware of the rounds when the other is chosen. We consider two variants of our model with different information structures. In the first, each agent knows both principals' algorithms, and the global round when (s)he arrives, \emph{but not} the rewards of the previous agents. In the second, agents only make decisions based on th rewards of previous agents. Concretely, each of the two principals has a \emph{reputation score}, and each agent's choice is driven by these two numbers. The reputation score is simply a sliding window average: an average reward of the last $M$ agents that chose this firm.

}
\xhdr{Some terminology.} The two principals are called ``Principal
1" and ``Principal 2".
%Principal $i\in \{1,2\}$ is also denoted \prin[i].
The algorithm of principal $i\in \{1,2\}$ is called ``algorithm $i$" and denoted
\alg[i]. The agent in global round $t$ is called ``agent $t$"; the
chosen principal is denoted $i_t$.

@@

\xhdr{Agents' response.}

Each agent $t$ chooses principal $i_t$ as follows: it chooses a distribution over the principals, and then draws independently from this distribution. Let $p_t$ be the probability of choosing principal $1$ according to this distribution. Below we specify $p_t$; we need to be careful so as to avoid a circular definition.

\gaedit{

The form of $p_t$ depends on the information structure that we consider.
In the first, where the agents only know the global round and the principals' algorithms, it is defined as follows.

}

@@

\gaedit{
In the second information structure we consider, at a given time $t$ each principal $i \in \{1, 2 \}$ has a reputation score denoted as $\REP_i (t)$. In this case we have that the agent's responses take an analogous form as in the first information structure:
\begin{align}
p_t = \respF\left(\; \REP_1(t) - \REP_2(t) \;\right).
\end{align}
}
\xhdr{Response functions.}
We use the response function $\respF$ \gaedit{to characterize the decision rule of the agents in our model}. We assume that $\respF$ is monotonically non-decreasing, is larger than $\tfrac12$ on the interval $(0,1]$, and smaller than $\tfrac12$ on the interval $[-1,0)$. Beyond that, we consider three specific models (see \reffig{fig:response-functions}):

@@

We will emphasize several specific algorithms or classes thereof. In the second half of the paper we will :

\begin{itemize}
\item ``smart" MAB algorithms that combine exploration and exploitation, such as UCB1 \cite{bandits-ucb1} and \gaedit{Thompson Sampling [citation for TS]}. These algorithms achieve
        $\BIR(n) \leq  \tilde{O}(n^{-1/2})$
    for all priors and all (or all but a very few) steps $n$. This bound is known to be tight for any fixed $n$.
    \footnote{This follows from the lower-bound analysis in \cite{bandits-exp3}.}

@@

When we consider the analytical model, we focus on MAB algorithms such that $\BIR(n)$ is non-increasing; we call such algorithms \emph{monotone}. While some reasonable MAB algorithms may occasionally violate monotonicity, they can usually be easily modified so that monotonicity violations either vanish altogether, or only occur at very specific rounds (so that agents are extremely unlikely to exploit them in practice).

@@

\gaedit{When we move to the variant of the model where we consider numerical simulations, we will pick a representative algorithm from each class of algorithms discussed here as our main interest is in qualitative differences between the different classes. We will utilize Thompson Sampling (\TS) from the ``smart" MAB algorithms, $\eps$-greedy (\DEG)\footnote{Throughout, we fix $\eps = 0.05$. Our pilot experiments showed that different $\eps$ did not qualitatively change the results.} from the ``naive" algorithms, and \DynGreedy (\DG). For ease of comparison, all three algorithms are parameterized with the same fake prior: namely, the mean reward of each arm is drawn independently from a $\Beta(1,1)$ distribution. Recall that Beta priors with 0-1 rewards form a conjugate family, which allows for simple posterior updates.
}

@@

\gaedit{
\xhdr{Varying Competition}

In both variants of the model that we consider, we have several levers for varying the nature of competition. In the strictest version of the model,

In some of our experiments, one firm is the ``incumbent" who enters the market before the other (``late entrant"), and therefore enjoys a \emph{temporary monopoly}. Formally, the incumbent enjoys additional $X$ rounds of the ``warm start". We treat $X$ as an exogenous element of the model, and study the consequences for a fixed $X$.

}

### 2019-08-18 journal paper/journal_main.tex
first pass through intro and model section

% notation
%%% advanced notation
\newcommand{\OPT}{\term{OPT}}
\newcommand{\rew}{\term{rew}}  % Bayesian-expected reward after n local rounds
\newcommand{\PMR}{\term{PMR}} % posterior mean reward
\newcommand{\support}{\term{support}}

@@

\begin{document}
\title[Competing Bandits]
{Competing Bandits}

@@

\begin{abstract}
  We empirically study the interplay between \textit{exploration} and
  \textit{competition}. Systems that learn from interactions with
  users often engage in \emph{exploration}: making potentially
  suboptimal decisions in order to acquire new information for future
  decisions. However, when multiple systems are competing for
    the same market of users, exploration may hurt a system's
    reputation in the near term, with adverse competitive effects. In particular, a system may enter a ``death spiral", when the short-term reputation cost decreases
    the number of users for the system to learn from, which degrades the
    system's performance relative to competition and further decreases
    the market share.

We ask whether better exploration algorithms are incentivized under competition. We run extensive numerical experiments in a stylized duopoly model in which two firms deploy multi-armed bandit algorithms and compete for myopic users.  We find that duopoly and monopoly tend to favor a primitive ``greedy algorithm" that does not explore and leads to low consumer welfare, whereas a temporary monopoly (a duopoly with an early entrant) may incentivize better bandit algorithms and lead to higher consumer welfare. Our findings shed light on the first-mover advantage in the digital economy by exploring the role that data can play as a barrier to entry in online markets.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
\subfile{itcs18paper/sec-intro}

@@

\section{Our model and preliminaries}
\label{sec:model}
\subfile{itcs18paper/sec-model}

@@

\section{Economic implications}
\label{sec:welfare}
\subfile{itcs18paper/sec-welfare}

\subfile{ec19paper/content/introduction}

\subfile{ec19paper/content/model}

@@

\subfile{ec19paper/content/non_greedy_choice}

\subfile{ec19paper/content/conclusion}

>>>>

% notation
%%% advanced notation
\newcommand{\OPT}{\term{OPT}}
\newcommand{\rew}{\term{rew}}  % Bayesian-expected reward after n local rounds
\newcommand{\REP}{\term{REP}} % posterior mean reward
\newcommand{\PMR}{\term{PMR}} % posterior mean reward
\newcommand{\support}{\term{support}}

@@

\begin{document}
\title[Competing Bandits: \\ The Perils of Exploration under Competition]
{Competing Bandits: \\ The Perils of Exploration under Competition}

@@

\begin{abstract}
 Abstract goes here
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
\subfile{content/sec-intro}

@@

\section{Our model and preliminaries}
\label{sec:model}
\subfile{content/sec-model}

@@

%\section{Economic implications}
%\label{sec:welfare}
%\subfile{itcs18paper/sec-welfare}

@@

%\subfile{ec19paper/content/non_greedy_choice}
